How to select predictive models for causal inference?

### 2023-11-16 how_to_select_causal_models.tex
wip: get back alst version of text

% declared functions and additional packages
\usepackage{caption}
\usepackage{subcaption}
\usepackage{threeparttable}
\usepackage{times}
\usepackage{hyperref}
\def\mycitecolor{green!50!black}
\def\mylinkcolor{red!60!black}
\hypersetup{ colorlinks=true, linkcolor=\mylinkcolor, citecolor=\mycitecolor }
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb} %mathbb packages
\usepackage{siunitx}
\usepackage{algorithm, algorithmicx, algpseudocode}
\usepackage{multirow}
\usepackage[sorting=nyt,style=numeric, natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
\DeclareNameAlias{default}{last-first}
\usepackage{placeins}
% associated arxiv class
\usepackage{arxiv}
\usepackage{multicol}
\setlength{\columnsep}{1cm}
\usepackage{authblk}% for author affiliations
\usepackage[dvipsnames]{xcolor}

\usepackage{comment}
%\usepackage{snaptodo}

@@

% \snaptodoset{block rise=1em}
% \snaptodoset{margin block/.style={font=\tiny}} \newcommand{\gv}[1]{%
%     \snaptodo[margin block/.append style=green!50!black]{%
% 	\sloppy\textbf{Gael}: #1}}

% \newcommand{\md}[1]{\snaptodo[margin block/.append style=blue]{\textbf{Matt}:
% #1}} \newcommand{\idea}[1]{\snaptodo[margin block/.append
% style=red]{\textbf{Idea}: #1}}

@@

\begin{abstract}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention,
    a causal-inference task. The increasing complexity of
    health data has opened the door to a plethora of models, but also the
    Pandora box of model selection: which of these models yield the most
    valid causal estimates?
    Here we highlight that classic machine-learning model selection
    does not select the best outcome models for causal inference.
    Indeed, causal model selection should control both outcome errors
    for each individual, treated or not
    treated, whereas only one outcome is observed.
    Theoretically, simple risks used in machine learning do not control
    causal effects when treated and non-treated population differ too much.
    More elaborate risks build proxies of the causal error using
    ``nuisance'' re-weighting to compute it on the observed data. But
    does computing these nuisance adds noise to model selection? Drawing
    from an extensive empirical study, we outline a good causal
    model-selection procedure: using the so-called
    $R\text{-risk}$; using flexible estimators to
    compute the nuisance models on the train set; and splitting out 10\% of
    the data to compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

@@

    \subsection{Extending prediction to prescription requires causal model selection}

    % Framing:
    % - Increasing rich observational, ``real-life'' data.
    %   - routinely collected health care
    %   - convenient to build prognostic model (eg using claims or
    %   administrative data
    %   \cite{desai2020comparison} ) \cite{yurkovich2015systematic},
    %   electronic health records \cite{fontana2019can}, encompassing notes and questionnaires
    %   \cite{horng2017creating,simon2018predicting}
    % - Motivates building increasingly complex predictive model, such as
    %   prognostic models
    % - Plethora of predictive models
    % - From prediction to prescription There is often a desire to reason on intervention => causal inference is necessary

    Increasingly rich data drives new predictive models. In health, new risks
    or prognostic models leverage routinely-collected data, sometimes with
    machine learning \cite{mooney2018bigdata}: predicting morbidity-related
    outcomes from administrative data \cite{yurkovich2015systematic}, heart
    failure from claims \cite{desai2020comparison}, sepsis from clinical
    records \cite{horng2017creating}, suicide attempts from patient records
    and questionnaires \cite{simon2018predicting}... Data may be difficult to
    control and model, but claims of accurate prediction can be established
    on left-out data
    \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
    Given a model predicting of an outcome of interest, it is tempting to use
    it to guide decisions: will an individual benefit or not from an
    intervention such as surgery \cite{fontana2019can}? This is a
    causal-inference task and principled approaches can build
    on contrasting the
    prediction of the outcome with and without the treatment
    \cite{snowden_implementation_2011,blakely2020reflection}.

@@

    Outcome modeling is an integral part of the causal modeling toolkit,
    under the names of
    G-computation, G-formula \cite{robins_role_1986}, Q-model
    \cite{snowden_implementation_2011}, or conditional mean
    regression \cite{wendling_comparing_2018}. On observational data, causal
    inference of treatment effects is brittle to un-accounted for
    confounding, but it can assess
    effectiveness and safety in real-word practice
    \cite{black1996we,hernan_methods_2021} or off-label drug usage
    \cite{radley2006off} for potential repurposing
    \cite{hurle2013computational,dudley2011exploiting}. Epidemiology has
    historically focused on methods that model treatment assignment
    \cite{austin_moving_2015,grose_use_2020}, based on the propensity score
    \cite{rosenbaum_central_1983} but recent empirical results
    \cite{wendling_comparing_2018,dorie_automated_2019} show benefits of
    outcome modeling to estimate average treatment effect (ATE). A major
    benefit of using outcome modeling for causal inference is that these
    methods naturally go beyond average effects,
    estimating individualized or conditional average treatment effects
    (CATE), important for precision medicine.
    %
    For this purpose, such methods are also invaluable on randomized trials
    \cite{su2018random,lamont2018identification,hoogland2021tutorial}.

    %\idea{There is a variety of models for g-estimation}
    Recent developments have seen the multiplication of predictive modeling
    methods. Leaving aside the, overwhelming, machine-learning literature,
    even methods specifically designed for causal
    inference are numerous: Bayesian Additive Regression Trees
    \cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
    \cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
    \cite{powers_methods_2018}, causal multivariate adaptive regression
    splines (MARS) \cite{powers_methods_2018}, random forests
    \cite{wager_estimation_2018, athey_generalized_2019},
    Meta-learners \cite{kunzel_metalearners_2019}, R-learners
    \cite{nie_quasioracle_2017}, Doubly robust estimation
    \cite{chernozhukov_double_2018}...
    %\idea{The variety of estimators calls for model selection procedures}
    The wide variety of methods leaves the applied researcher with the difficult choice
    of selecting between different estimators based on the data at hand.
    %
    Indeed, estimates may vary markedly when using different
    models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
    large variations obtained across four different outcome estimators on
    2016 semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
    such as random forests are doing well in most settings except
    for setups where treated and untreated populations differ markedly in
    which case a simple linear model (ridge) is to be preferred.
    However a different choice of hyper-parameters
    (max depth= 2) for random forests yield the poorest performances.
    A simple rule of thumb such as preferring more flexible models does not work in
    general; model selection is necessary.

@@

    \begin{figure}[b!]
        \begin{minipage}{.4\linewidth}
            \caption{\textbf{Different outcome models lead to different
                    estimation error on the Average Treatment Effects},
                with six different outcome
                models on 77 classic simulations where the true causal effect is
                known \cite{dorie_automated_2019}. The models are random forests, ridge regression with and without
                interaction with the treatment,
                (hyper-parameters detailed in Appendix
                \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
                plotted as a function of increasing difference between treated and
                untreated population --detailed in
                \autoref{subsec:measuring_overlap}.  \\[1ex]
                There is no systematic best performer: choosing
                the best model among a family of candidate estimators is
                important.
            }\label{fig:acic_2016_ate_heterogeneity}%
        \end{minipage}\hfill%
        \begin{minipage}{.6\linewidth}
            \centering
            \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
        \end{minipage}%
    \end{figure}

    Standard practices to select models in predictive settings rely on
    cross-validation on the error on the outcome
    \cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
    will see, these model-selection practices may not pick the best models
    for causal inference, as they can be misled by inhomogeneities due to
    treatment allocation.
    %
    Given complex, potentially noisy, data, which model is to be most trusted to
    yield valid causal estimates? Because there is no single learner that performs
    best on all data sets, there is a pressing need for clear guidelines to select
    outcome models for causal inference.

@@

    In this paper, we study \textit{model selection procedures} with a focus
    on practical settings: \textit{finite samples} settings and without
    \textit{well-specification} assumption. One question is whether
    model-selection procedures, that rely on data split, can estimate
    reliably enough the complex risks, theoretically motivated for causal
    inference. Indeed, these risks force of departure from standard model-selection
    procedures as they come with more quantities to estimate, which may
    bring additional variance, leading to worse model selection.

    We first give a simple illustration of the problem of causal model
    selection and quickly review the important prior art. Then, in Section
    \ref{sec:framework}, we set causal model selection in the
    \emph{potential outcome} framework and detail the causal risks and
    model-selection procedure. Section \ref{sec:theory} gives our theoretical
    result. In section \ref{sec:empirical_study}, we run a thorough empirical
    study, with many different settings covered. Finally, we comment our
    findings in Section \ref{sec:discussion}.
    %
    Results outline how to best select outcome models for causal
    inference with an adapted
    cross-validation that estimate the so-called $R\text{-risk}$.
    The $R\text{-risk}$
    modulates observed prediction error to compensate for systematic
    differences between treated and non-treated individuals. It
    relies on the two \emph{nuisance} models,
    themselves estimated from data and thus imperfect; yet these
    imperfections do not undermine the benefit of the $R\text{-risk}$.

    % XXX: merge the two paragraphs, above and below

@@

    Using a predictor to reason on causal effects relies on contrasting the
    prediction of the outcome for a given individual with and without the
    treatment --as detailed in \autoref{sec:neyman_rubin}.
    %
    Given various predictors of the outcome, we are interested in
    selecting those that estimate best the treatment effect.
    %
    Standard predictive modeling or machine-learning practice selects the
    best predictor, \emph{ie} the one that minimizes the expected error.
    However, the best predictor may not be the best model to reason about
    causal effects of an intervention, as we illustrate below.

@@

    \section{Formal setting of causal inference and model selection}\label{sec:framework}

    \subsection{The Neyman-Rubin Potential Outcomes framework}%
    \label{sec:neyman_rubin}%

@@

    \paragraph{Oracle Bayes predictor}\label{remark:bayes_oracle} Consider the case
    where we have access to the oracle Bayes predictor for the outcome ie.~$f(x,
        a)=\mu(x, a)$, then all risks are equivalent up to the residual variance:
    \begin{equation}
        \tau\text{-risk}(\mu) = \mathbb E_{X\sim p(X)}[(\tau(X) - \tau_{\mu}(X))^2] = 0
    \end{equation}
    \begin{equation}
        \mu\text{-risk}(\mu) = \mathbb E_{(Y, X, A) \sim p(Y;X;A)}[\big( Y - \mu_A(X)\big)^2] = \int_{\mathcal X, \mathcal A}
        \,\varepsilon(x,a)^2 p(a \mid x) \,p(x) \,dx\,da  \leq \sigma_B^{2}(0) + \sigma_B^{2}(1)
    \end{equation}
    \hspace*{0.5em}
    \begin{equation}
        \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)  \quad \text{follows from Lemma \ref{apd:proofs:mu_risk_ipw_link_mu}}
        %\notag
    \end{equation}
    \hspace*{0.5em}
    \begin{equation}
        R\text{-risk}(\mu) = \tilde{\sigma}_B^{2}(0) + \tilde{\sigma}_B^{2}(1)
        \leq \sigma_B^{2}(0) + \sigma_B^{2}(1) \quad \text{follows directly from Proposition \ref{theory:prop:r_risk_rewrite}}%\notag
    \end{equation}
    Thus, differences between causal risks only matter in finite sample regimes.
    Universally consistent learners converge to the Bayes risk in asymptotic
    regimes, making all model selection risks equivalent. However, in practice
    choices must be made in non-asymptotic regimes.

>>>>

% declared functions and additional packages
\usepackage{caption}
\usepackage{subcaption}
\usepackage{threeparttable}
\usepackage{times}
\usepackage{hyperref}
\def\mycitecolor{green!50!black}
\def\mylinkcolor{red!60!black}
\hypersetup{ colorlinks=true, linkcolor=\mylinkcolor, citecolor=\mycitecolor }
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb} %mathbb packages
\usepackage{siunitx}
\usepackage{algorithm, algorithmicx, algpseudocode}
\usepackage{multirow}
\usepackage[sorting=nyt,style=numeric, natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
\DeclareNameAlias{default}{last-first}
\usepackage{placeins}
% associated arxiv class
\usepackage{arxiv}
\usepackage{multicol}
\setlength{\columnsep}{1cm}
\usepackage{wrapfig}
%\usepackage{wraptable}
\usepackage{authblk}% for author affiliations
\usepackage[dvipsnames]{xcolor}

\usepackage{comment}
\setcounter{secnumdepth}{0}

@@

@@

\begin{abstract}
    \textbf{Background:}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention,
    a causal-inference task. The increasing complexity of
    health data has opened the door to a plethora of models, but also the
    Pandora box of model selection: which of these models yield the most
    valid causal estimates? Causal model selection should control both outcome errors for each
    individual, treated or not treated, whereas only one outcome is observed.
    Theoretically, simple risks used in machine learning do not control causal
    effects when treated and non-treated population differ too much. \\
    \textbf{Methods:}
    More elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated datasets based on real
    covariates, we compare different metrics for causal model selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

@@

    \subsection{Extending prediction to prescription needs causality}

    Progress in machine learning brings predictive models to new health
    data \cite{beam2018big,rajkomar2019machine}, with the promise of
    precision medicine.
    Automated analysis of
    medical images is increasingly accurate, \emph{eg} for brain images,
    \cite{khojaste2022deep,zhang2019radiological} or mammography
    \cite{yala2019deep,shen2019deep,nassif2022breast}.
    New prognostic models leverage routinely-collected
    patient records
    \cite{mooney2018bigdata}: predicting heart
    failure from claims \cite{desai2020comparison}, suicide attempts
    from questionnaires \cite{simon2018predicting}...
    Clinical notes contain much prognostic information
    but require text modeling  \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
    Data may be difficult to
    control and model, but the accuracy of the prediction can be verified
    on left-out data
    \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
    Given a model predicting a health outcome, precision medicine
    would like it to guide decisions: will an individual benefit from an
    intervention such as surgery \cite{fontana2019can}?
    Contrasting predictions with and without the treatment gives an answer,
    but statistical validity requires causal inference
    \cite{snowden_implementation_2011,blakely2020reflection}.

@@

    There is a rich causal-inference statistical literature, which bridges to
    predictive modeling via so-called \emph{outcome models}, or G-computation,
    G-formula \cite{robins_role_1986}, Q-model \cite{snowden_implementation_2011},
    conditional mean regression \cite{wendling_comparing_2018}. A central challenge
    of inference of treatment effects is that of confounding: spurious associations
    between treatment allocation and baseline health, \emph{eg} only prescribing a
    drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
    Controlled allocation of treatment, as in randomized controlled trials (RCTs),
    alleviate this concern. Yet most machine-learning models are trained on
    \emph{observational} data, close to real-word practice
    \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
    Causal inference has been central to epidemiology, typically with methods that
    model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
    propensity scores \cite{rosenbaum_central_1983}. Recent empirical results
    \cite{wendling_comparing_2018,dorie_automated_2019} show benefits of outcome
    modeling to estimate average treatment effects. Maybe a greater benefit is
    that these methods naturally go beyond average effects, estimating
    individualized or conditional average treatment effects (CATE), central to
    precision medicine.
    %
    For this purpose, such methods are also invaluable on randomized trials
    \cite{su2018random,lamont2018identification,hoogland2021tutorial}.

    %\idea{There is a variety of models for g-estimation}
    There is an explosion of outcome modeling or machine learning methods.
    For instance, many deep-learning methods have been developed for medical
    image analysis \cite{shen2017deep,monshi2020deep}.
    Even outcome-modeling methods specifically designed for causal
    inference are numerous: Bayesian Additive Regression Trees
    \cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
    \cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
    \cite{powers_methods_2018}, causal multivariate adaptive regression
    splines \cite{powers_methods_2018}, random forests
    \cite{wager_estimation_2018, athey_generalized_2019},
    Meta-learners \cite{kunzel_metalearners_2019}, R-learners
    \cite{nie_quasioracle_2017}, Doubly robust estimation
    \cite{chernozhukov_double_2018}...
    %\idea{The variety of estimators calls for model selection procedures}
    The wide variety of methods raises the problem
    of selecting between different estimators based on the data at hand.
    %
    Indeed, estimates of treatment effects can vary markedly across different
    predictive models. For instance, Figure shows
    large variations obtained across different outcome estimators on
    semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
    such as random forests are doing well in most settings except
    when treated and untreated populations differ noticeably, in
    which case a linear model (ridge) is to be preferred.
    However random forests with different hyper-parameters
    (max depth= 2) yield poor estimates.
    A simple rule of thumb such as preferring flexible models does not work in
    general; model selection is needed.

@@

    \begin{figure}
        \centering
        \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
        \caption{\textbf{Different outcome models lead to different
                estimation errors on the Average Treatment Effects},
            on 77 classic simulations with known true causal effect
            \cite{dorie_automated_2019}. The different models are ridge regression
            and random forests with different hyper-parameters
            (details
            \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
            plotted as a function of increasing difference between treated and
            untreated population --see
            \autoref{subsec:measuring_overlap}.
            There is no systematic best performer; data-driven model
            selection is important.
            \label{fig:acic_2016_ate_heterogeneity}%
        }
    \end{figure}

    Standard practices to select models in predictive settings rely on
    the error on the outcome
    \cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
    will see, these practices may not pick the best models
    for causal inference, as they can be misled by inhomogeneities due to
    treatment allocation.
    %
    Given complex, potentially noisy, data, which model is to be most trusted to
    yield valid causal estimates? As no single learner performs
    best on all data sets, there is a pressing need for clear guidelines to select
    outcome models for causal inference.

@@

    In this paper, we study \textit{model selection procedures}
    in practical settings: \textit{finite samples} settings and without
    \textit{well-specification} assumption. Asymptotic causal-inference
    theory calls for complex risks, but a practical question is
    whether model-selection procedures, that rely on data split, can estimate
    these risks reliably enough. Indeed, they
    come with more quantities to estimate, which may
    bring additional variance, leading to worse model selection.

    We first illustrate the problem of causal model
    selection and briefly review prior art. Then, Section
    \ref{sec:framework} sets causal model selection in the
    \emph{potential outcome} framework and details the causal risks and
    model-selection procedure. Section \ref{sec:theory} gives theoretical
    results. Section \ref{sec:empirical_study} details a thorough empirical
    study, covering many different settings. Finally, Section \ref{sec:discussion}
    discusses the findings.
    %
    Results outline how to best select outcome models for causal
    inference with an adapted
    cross-validation to estimate the so-called $R\text{-risk}$.
    This risk compensates for systematic
    differences between treated and non-treated individuals using
    two \emph{nuisance} models,
    themselves estimated from data and thus imperfect; yet these
    imperfections do not undermine the $R\text{-risk}$.

@@

    Using a predictor to reason on causal effects relies on contrasting the
    prediction of the outcome for a given individual with and without the
    treatment --as detailed in the Methods \autoref{sec:neyman_rubin}.
    %
    Given various predictors of the outcome, we are interested in
    selecting those that estimate best the treatment effect.
    %
    Standard predictive modeling or machine-learning practice selects the
    best predictor, \emph{ie} the one that minimizes the expected error.
    However, the best predictor may not be the best model to reason about
    causal effects of an intervention, as we illustrate below.

@@

    %\section{Formal setting of causal inference and model selection}\label{sec:framework}
    \section{Methods}\label{sec:framework}
    \subsection{The Neyman-Rubin Potential Outcomes framework}%
    \label{sec:neyman_rubin}%

@@

    \paragraph{Oracle Bayes predictor}\label{remark:bayes_oracle} Consider the case
    where we have access to the oracle Bayes predictor for the outcome ie.~$f(x,
        a)=\mu(x, a)$, then all risks are equivalent up to the residual variance:
    \begin{equation}
        \tau\text{-risk}(\mu) = \mathbb E_{X\sim p(X)}[(\tau(X) - \tau_{\mu}(X))^2] = 0
    \end{equation}
    \begin{align}
        \mu\text{-risk}(\mu) & = \mathbb E_{(Y, X, A) \sim p(Y;X;A)}[\big( Y - \mu_A(X)\big)^2] \\
                             & = \int_{\mathcal X, \mathcal A}
        \,\varepsilon(x,a)^2 p(a \mid x) \,p(x) \,dx\,da  \leq \sigma_B^{2}(0) + \sigma_B^{2}(1) \nonumber
    \end{align}
    \begin{equation}
        \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)  \quad \text{from Lemma \ref{apd:proofs:mu_risk_ipw_link_mu}}
        %\notag
    \end{equation}
    \begin{multline}
        R\text{-risk}(\mu) = \tilde{\sigma}_B^{2}(0) + \tilde{\sigma}_B^{2}(1)
        \leq \sigma_B^{2}(0) + \sigma_B^{2}(1)  \\  \text{from Proposition \ref{theory:prop:r_risk_rewrite}}%\notag
    \end{multline}
    Thus, differences between causal risks only matter in finite sample regimes.
    Universally consistent learners converge to the Bayes risk in asymptotic
    regimes, making all model selection risks equivalent. However, in practice
    choices must be made in non-asymptotic regimes.
